---
title: ".b[Dynamic regression models: Further thoughts]"
subtitle: ".b[.green[EC 361--001]]"
author: "Prof. Santetti"
date: "Spring 2024"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---


```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork, janitor, tsibble, lubridate, fpp3, MetBrewer)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 8,
  fig.width = 12.5,
  warning = F,
  dev = "svg",
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())

```

```{R, colors, include = F}
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#E02C05"
blue <- "#2b59c3"
green <- "#0FDA6D"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
```






# Materials

<br>

.b[Required readings]:

<br>


  - [`Hyndman & Athanasopoulos, ch. 10`](https://otexts.com/fpp3/dynamic.html)
  
    - sections 10.3; 10.5.
    


---
class: inverse, middle

# Motivation


---

# Motivation

<br><br><br>


After studying the .hi-orange[essentials] of *dynamic regression* models, let us wrap up this topic with a few more .hi-red[practical issues].

---
class: inverse, middle

# Harmonic regression with ARIMA errors

---

# Harmonic regression with ARIMA errors


We have already seen that .hi[harmonic regression] (i.e., using *Fourier* terms) performs well for modeling .hi-red[seasonality].

--

  - Especially *long* seasonal periods (*weekly* and *monthly* seasonal data, for example).
  
--

<br>

One .hi-red[limitation], however, is that harmonic regression assumes that the seasonal component is .hi-blue[fixed] over time (i.e., its variance is constant).

--

With this in mind, harmonic regression is a *useful tool* for seasonal time series.

<br>

--

*Fourier terms* can be used .hi-slate[along with] *ARIMA errors* to capture several important dynamics of the time series at hand.

---


# Harmonic regression with ARIMA errors

```{r, echo=F}
air <- read_csv("air_passengers.csv")


air_ts <- air |> 
  mutate(date = yearmonth(date)) |> 
  as_tsibble(index = date)

air_ts |> 
  ggplot(aes(x = date, y = passengers)) +
  geom_line() +
  labs(title = "International airline passengers",
       subtitle = "Jan 1949 – Dec 1960",
       caption = "Source: Brown (1962).",
       x = "",
       y = "Thousands") +
  easy_y_axis_title_size(14) +
  easy_plot_caption_size(14)
```


---


# Harmonic regression with ARIMA errors

- Let us start a forecasting exercise for the .hi-slate[air passengers] data set with a *straightforward* regression model. 

--

<br>

```{r}
air_tslm <- air_ts |> 
  model(reg = TSLM(log(passengers) ~ trend() + season()))
```



---
class: clear



.smallest[
```{r}
air_tslm |> 
  report()
```
]


---
class: clear


.smaller[
```{r, fig.height=7.5}
air_tslm |> 
  augment() |> 
  gg_tsdisplay(.innov, plot_type = "partial")
```
]

---


# Harmonic regression with ARIMA errors

- Testing for .hi-slate[residual serial correlation]:

```{r}

air_tslm |> 
  augment() |> 
  features(.innov, ljung_box, lag = 2 * 12)
```


<br>

What do we conclude?
---


# Harmonic regression with ARIMA errors

Moving on to a model with .hi-blue[ARIMA errors]:

<br>

```{r}
air_arima <- air_ts |> 
  model(arima_errors = ARIMA(log(passengers) ~ trend() + season() + PDQ(0, 0, 0))) # setting PDQ() with zeros, as
                                                                                   # the "season()" function is
                                                                                   # taking care of seasonality.
```


---
class: clear

.smallest[
```{r}
air_arima |>
  report()
```

]



---
class: clear


.smaller[
```{r, fig.height=7.5}
air_arima |> 
  augment() |> 
  gg_tsdisplay(.innov, plot_type = "partial")
```
]


---


# Harmonic regression with ARIMA errors

- Testing for .hi-slate[residual serial correlation]:

```{r}
air_arima |> 
  augment() |> 
  features(.innov, ljung_box, lag = 2 * 12)
```


<br>

What do we conclude?

---


# Harmonic regression with ARIMA errors

In case we forecast the time series at hand based on a model *with residual serial correlation*, .hi-blue[prediction intervals] will be .hi-red[unreliable].

--

```{r}
air_arima |> 
  forecast(h = 24) |> 
  head(6)
```


---
class: clear

.smaller[
```{r}
air_arima |> forecast(h = 24) |> autoplot(air_ts, level = 95, linewidth = .8) + labs(x = "")
```
]


---


# Harmonic regression with ARIMA errors

So far, we have not been successful in coming up with a *reliable* .hi-slate[regression forecast model] for the air passengers data set.

--

An *alternative* for modeling seasonality is using .hi-red[harmonic regression] .hi-slate[with ARIMA errors].

--

<br>


```{r}
air_harmonic_fit <- model(air_ts,
              K1 = ARIMA(log(passengers) ~ fourier(K = 1) + PDQ(0, 0, 0)),
              K2 = ARIMA(log(passengers) ~ fourier(K = 2) + PDQ(0, 0, 0)),
              K3 = ARIMA(log(passengers) ~ fourier(K = 3) + PDQ(0, 0, 0)),
              K4 = ARIMA(log(passengers) ~ fourier(K = 4) + PDQ(0, 0, 0)),
              K5 = ARIMA(log(passengers) ~ fourier(K = 5) + PDQ(0, 0, 0)),
              K6 = ARIMA(log(passengers) ~ fourier(K = 6) + PDQ(0, 0, 0)))

```

<br>

Recall that *K*, the number of Fourier sine and cosine pairs, can vary from *K = 1* up to *K = m/2*.


---
class: clear

```{r, echo=F, fig.height = 9, fig.width=15}
air_harmonic_fit |>
  forecast(h = 24) |>
  autoplot(air_ts, level = 95) +
  facet_wrap(vars(.model), ncol = 2) +
  guides(colour = "none", fill = "none", level = "none") +
  geom_label(aes(x = yearmonth("1953 Jan"), y = 700, label = paste0("AICc = ", format(AICc))),
    data = glance(air_harmonic_fit)) +
  labs(title= "U.S. air passengers: 24-month ahead forecast",
       y="Thousands",
       x = "")

```






---
class: clear


.smallest[
```{r, fig.height=7}

air_harmonic_fit |> 
  select(K6) |> 
  augment() |> 
  gg_tsdisplay(.innov, plot_type = "partial")
```
]


---


# Harmonic regression with ARIMA errors



.pull-left[
.small[
```{r}
air_harmonic_fit |> 
  select(K6)
```


```{r}
air_harmonic_fit |> 
  select(K5)
```
]]


.pull-right[
.small[
```{r}
air_harmonic_fit |> 
  select(K6) |> 
  augment() |> 
  features(.innov, ljung_box, lag = 2 * 12, dof = 2)
```


```{r}
air_harmonic_fit |> 
  select(K5) |> 
  augment() |> 
  features(.innov, ljung_box, lag = 2 * 12, dof = 2)
```
]]




---


# Harmonic regression with ARIMA errors

```{r, echo=F}
air_harmonic_fit |> 
  select(K5) |> 
  forecast(h = 24) |> 
  autoplot(air_ts, level = 95, color = "#3c3b6e", linewidth = .8) +
  labs(title = "24-month ahead forecast (Dynamic harmonic regression, K = 5)",
       subtitle = "U.S. air passengers data",
       x = "",
       y = "Thousands") +
  easy_y_axis_title_size(13)
```


---
class: inverse, middle

# Scenario-based forecasting


---


# Scenario-based forecasting

When using .hi-slate[regression] models for forecasting, the time series we are interested in forecasting is a .hi[function] of one or more .hi-blue[predictor variables].

--

<br>

Excluding variables we .hi-blue[do know] their future values (e.g., *trend* and *seasonal dummy* variables), we must .hi-slate[make assumptions] about other predictors for which we .hi-red[do not have] information.

--

<br>

This allows us to assume different .hi-slate[scenarios] for the *predictor variables* that are of interest.

--

<br>

Let us come back to our .hi-red[Phillips curve] example.



---


# Scenario-based forecasting

```{r, echo=F, message=F, warning=F}
dat <- read_csv("ps1_data.csv")


phillips_ts <- dat |> 
  mutate(date = floor_date(period, unit = "year")) |> 
  group_by(date) |> 
  summarize(infrate = mean(infrate),
            unrate = mean(unrate)) 


phillips_ts <- phillips_ts |> 
  mutate(date = year(date)) |> 
  as_tsibble(index = date)

phillips_ts <- phillips_ts |> 
  mutate(delta_infrate = difference(infrate, lag = 1))

```






```{r}

phillips_arima <- phillips_ts |> 
  model(arima_reg = ARIMA(delta_infrate ~ unrate))
```


--

<br>

We will assume .b[3] possible .hi-blue[forecasting scenarios]:

<br>

1. The unemployment rate follows its historical *average value* (5.9%);

2. The unemployment rate *increases* to 10%;

3. The unemployment rate follows its *current value* of 3.36%.

---


# Scenario-based forecasting

<br>


```{r}
future_scenarios <- scenarios(average = new_data(phillips_ts, n = 5) |> 
                                mutate(unrate = mean(phillips_ts$unrate)),
                              pessimistic = new_data(phillips_ts, n = 5) |> 
                                mutate(unrate = 10),
                              optimistic = new_data(phillips_ts, n = 5) |> 
                                mutate(unrate = 3.36))
```


--

<br>


```{r}
phillips_scen_fc <- phillips_arima |> 
  forecast(h = 5, new_data = future_scenarios)
```


---
class: clear

.smallest[
```{r, fig.height=7.5}
phillips_ts |> 
  autoplot(delta_infrate) +
  autolayer(phillips_scen_fc |> filter(.scenario == "average"), level = 95, color = "#4682b4", linewidth = .8) +
  labs(title = "Average scenario")
```
]


---
class: clear

.smallest[
```{r, fig.height=7.5}
phillips_ts |> 
  autoplot(delta_infrate) +
  autolayer(phillips_scen_fc |> filter(.scenario == "pessimistic"), level = 95, color = "#800020", linewidth = .8) +
  labs(title = "Pessimistic scenario")
```
]


---
class: clear

.smallest[
```{r, fig.height=7.5}
phillips_ts |> 
  autoplot(delta_infrate) +
  autolayer(phillips_scen_fc |> filter(.scenario == "optimistic"), level = 95, color = "#228b22", linewidth = .8) +
  labs(title = "Optimistic scenario")
```
]

---
class: clear

```{r, echo=F, fig.height=9}
phillips_ts |> 
  autoplot(delta_infrate) +
  autolayer(phillips_scen_fc, level = NULL, linewidth = 0.8) +
  scale_color_manual(values= c("#4682b4", "#800020", "#228b22")) +
  labs(title = "All 3 scenarios")
```


---
layout: false
class: inverse, middle

# Next time: Applications

---
exclude: true
